---
title: "Creating mock datasets"
author: "Fleur Ballet"
output: html_document
date: "2024-05-27"
---

```{r setup, include = FALSE}
## Options for Rmarkdown compilation
knitr::opts_chunk$set(fig.width = 7,
                      fig.height = 5,
                      fig.align = "center",
                      out.width = "70%",
                      message = FALSE,
                      collapse = TRUE,
                      crop = NULL, warning = FALSE)
```

```{r}
## Single-cell and protoeomics processing tools
library("scpdata")
library("scp")
## Visualization and manipulation tools
library("ggplot2")
library("dplyr")
library("patchwork")

library("impute")
library("scater")
library("msqrob2")
library("tibble")
```

# Functions

Define different functions for adapting parts of the existing dataset:
- to take subsets of the dataset with only one of the cell sample types.
- to make another dataset where the sample types are randomised.
- to take a subset of the dataset where we only keep a one sample type per run.
- to make a dataset with data from one sample type and randomly split into two fake types.
- to make a dataset with data from one sample type and per run split into two fake types.
- to insert an effect in another mock dataset.

```{r}
# take the subset of the sce where the variable takes the given value
subset_sce <- function(sce, variable, value) {
  new_sce <- sce
  new_sce <- new_sce[,colData(new_sce)[,variable] == value]
  return (new_sce)
}

# make a new sce where the values of the given variable are randomised
randomise_variable_sce <- function(sce, variable) {
  set.seed(123)
  new_sce <- sce
  values <- unique(colData(new_sce)[, variable])
  colData(new_sce)[variable] <- sample(values, ncol(new_sce), replace = TRUE)
  return (new_sce)
}

# take a subset of the sce so that in each run there are only samples with the same value for the given variable
subset_variable_per_run_sce <- function(sce, variable) {
  set.seed(123)
  sets <- unique(colData(sce)[["Set"]])
  for (set in sets) {
    set <- which(colData(sce)[["Set"]] == set)
    values <- colData(sce)[set, variable]
    new_value <- sample(values, 1)
    new_sce <- sce[,rownames(colData(sce)[,variable==new_value])]
  }
  return (new_sce)
}

# take the subset of the sce for which the given variable takes the given value,
# then randomly assign two new fake values of this variable to each entry
subset_and_split_sce <- function(sce, variable, value) {
  set.seed(123)
  new_sce <- subset_sce(sce, variable, value)
  values <- c(paste0(value,"type1"), paste0(value,"type2"))
  colData(new_sce)[variable] <- sample(values, ncol(new_sce), replace = TRUE)
  return (new_sce)
}
  
# take the subset of the sce for which the given variable takes the given value,
# then randomly assign two new fake values of this variable, 
# so that all entries from the same run take the same new value
subset_and_split_per_run_sce <- function(sce, variable, value) {
  set.seed(123)
  new_sce <- subset_sce(sce, variable, value)
  values <- c(paste0(value,"type1"), paste0(value,"type2"))
  sets <- unique(colData(new_sce)[["Set"]])
  for (set in sets) {
    set <- which(colData(new_sce)[["Set"]] == set)
    new_value <- sample(values, 1)
    colData(new_sce)[set, variable] <- rep(new_value, length(set))
  }
  return (new_sce)
}

# insert an effect in a mock sce by switching the intensities of the entries
# with the given value for the given variable
insert_effect <- function(sce, variable, value) {
  set.seed(123)
  new_sce <- sce
  samples <- which(colData(new_sce)[, variable] == value)
  for (sample in samples) {
    intensities <- assay(new_sce)[,sample]
    indices <- sample(1:length(intensities), floor(length(intensities)/5))
    intensities[indices] <- sample(intensities[indices])
    assay(new_sce)[,sample] <- intensities
  }
  return (new_sce)
}

# perform the given function with the given arguments on each sce in the list sces
# and return a list with the adapted sces
create_dataset_for_all <- function(func, sces, args) {
  new_sces <- list()
  for (name in names(sces)) {
    new_sces[[name]] <- do.call(func, c(sce = sces[[name]], args))
  }
  return (new_sces)
}
```


# Create new datasets

First load the data. This is a list of SingleCellExperiment-assays 
(as created after "minimal_preprocessing" and "preprocessing_continued").

```{r}
loadRData <- function(fileName){
    load(fileName)
    get(ls()[ls() != "fileName"])
}

sces <- loadRData("Data/leduc_sces.Rda")
```

Create the different new datasets with the previously defined functions, for each assay in sces.

```{r}
sces_subset_split <- create_dataset_for_all(func = subset_and_split_sce, sces, 
                                        args = list(variable="SampleType", value = "Monocyte"))
sces_subset_split_per_run <- create_dataset_for_all(func = subset_and_split_per_run_sce, sces, 
                                        args = list(variable="SampleType", value = "Monocyte"))
```

# Save results

```{r}
save(sces_subset_split, file = "Data/leduc_sces_subset_split.Rda")
save(sces_subset_split_per_run, file = "Data/leduc_sces_subset_split_per_run.Rda")
```

For the rest of the analyses, we will use "sces" (the real data),
and "sces_subset_split" and "sces_subset_split_per_run" (the mock data).