---
title: "Density estimation - combined groups"
author: "Fleur Ballet"
output: html_document
date: "2024-05-27"
---

```{r setup, include = FALSE}
## Options for Rmarkdown compilation
knitr::opts_chunk$set(fig.width = 7,
                      fig.height = 5,
                      fig.align = "center",
                      out.width = "70%",
                      message = FALSE,
                      collapse = TRUE,
                      crop = NULL, warning = FALSE)
```

In this script we implement the Poisson regression density estimation technique for all data combined.

Loading the packages:

```{r}
library("scp")
library("ggplot2")
library("dplyr")
library("patchwork")
library("car")
library("mgcv")
```

We load the data that was already preprocessed in the scripts "minimal_preprocessing" and
"preprocessing_continued". For the illustration of the methods we will work with the assay
"peptidesSweepImp" from the original data in "leduc_sces".

```{r}
# function to load the saved preprocessed assays
loadRData <- function(fileName){
    load(fileName)
    get(ls()[ls() != "fileName"])
}

sces <- loadRData("Data/leduc_sces.Rda")
sce <- sces[["peptidesSweepImp"]]
colData(sce)[["SampleType"]] <- ifelse(colData(sce)[["SampleType"]] == "Melanoma cell", 
                                       "Melanoma", "Monocyte") # rename for easier use
```

We define the function to discretise the intensity data contained in the assay:

```{r}
discretize <- function(data, n_bars) {
  # define the discretisation
  min <- min(data)
  max <- max(data)
  bar_width <- (max - min)/n_bars
  breaks <- seq(from = min, to = max, by = bar_width)
  mids <- seq(from = min+bar_width/2, length.out = length(breaks)-1, by = bar_width)
  # count the intensities in each interval
  counts <- sapply(1:(length(breaks)-1), 
                   function(b) sum(data > breaks[b] & data < breaks[b+1]))
  # store
  results <- list(n_bars, min, max, bar_width, breaks, mids, counts)
  names(results) <- c("n_bars", "min", "max", "bar_width", "breaks", "mids", "counts")
  return (results)
}
```

We discretise the intensities from the first peptide in the assay.

```{r}
intensities <- data.frame("intensities" = c(assay(sce[1,])))
discretized <- discretize(intensities, 40)
```

We define the data we need to fit the model: it contains the discretised sequence, the counts 
and the carrier. We will use a non-normalised uniform carrier. Then we can fit the desired model. 
We include the intensities up to the fourth degree for the GLM Poisson model and 
we also fit a smoothing GAM model.

```{r}
carrier <- rep(sum(discretized$counts)*(discretized$breaks[2]-discretized$breaks[1]),
               length(discretized$counts))

data <- data.frame("intensities" = discretized$mids,
                  "counts" = discretized$counts, "carrier" = carrier)
model <- glm(counts ~ intensities + I(intensities^2) + I(intensities^3) + I(intensities^4) +
               offset(log(carrier)), data, family = "poisson")
model2 <- gam(counts ~ s(intensities) + offset(log(carrier)), data, family = "poisson")
```
  
We can then plot the histogram of the counts and the predicted counts in each interval 
under the two models.

```{r}
ggplot(data) + geom_bar(aes(x = intensities, y = counts), stat = "identity") +
  geom_line(aes(x = intensities, y = model$fitted.values))
ggplot(data) + geom_bar(aes(x = intensities, y = counts), stat = "identity") +
  geom_line(aes(x = intensities, y = model2$fitted.values))
```

We now plot frequency histograms where we plot the predicted densities for a sequence
of 2000 new data points.

```{r}
newdata <- data.frame("intensities" = seq(discretized$min, discretized$max, length = 2000),
                      "carrier" = rep(1, length = 2000))
predictions <- data.frame("intensities" = newdata$intensities,
                   "predictions" = predict(model, newdata = newdata, type = "response"))

ggplot() + 
  geom_histogram(data = intensities, breaks = discretized$breaks,
                 mapping = aes(x = intensities, y = after_stat(density))) +
  geom_line(data = predictions, mapping = aes(x = intensities, y = predictions))
```

We want to create a plot that compares the densities predicted by the
special exponential family method and a kernel density method.

```{r}
data_sef <- data.frame("intensities" = discretized$mids,
                  "counts" = discretized$counts, "carrier" = carrier)
model_sef <- glm(counts ~ intensities + I(intensities^2) + I(intensities^3) + I(intensities^4),
             offset = log(carrier), data_sef, family = "poisson")
newdata_sef <- data.frame("intensities" = seq(discretized$min, discretized$max, length = 2000),
                      "carrier" = rep(1, length = 2000))
predictions_sef <- data.frame("intensities" = newdata_sef$intensities,
                   "predictions" = predict(model_sef, newdata = newdata_sef, type = "response"))

data_sef2 <- data.frame("intensities" = discretized$mids,
                  "counts" = discretized$counts, "carrier" = carrier)
model_sef2 <- glm(counts ~ intensities + I(intensities^2),
             offset = log(carrier), data_sef2, family = "poisson")
newdata_sef2 <- data.frame("intensities" = seq(discretized$min, discretized$max, length = 2000),
                      "carrier" = rep(1, length = 2000))
predictions_sef2 <- data.frame("intensities" = newdata_sef2$intensities,
                   "predictions" = predict(model_sef2, newdata = newdata_sef2, type = "response"))

data_gam <- data.frame("intensities" = discretized$mids,
                  "counts" = discretized$counts, "carrier" = carrier)
model_gam <- gam(counts ~ s(intensities), offset = log(carrier), data_gam, family = "poisson")
newdata_gam <- data.frame("intensities" = seq(discretized$min, discretized$max, length = 2000),
                      "carrier" = rep(1, length = 2000))
predictions_gam <- data.frame("intensities" = newdata_gam$intensities,
                   "predictions" = predict(model_gam, newdata = newdata_gam, type = "response"))

data_kde <- data.frame("intensities" = rep(discretized$mids, discretized$counts))
model_kde <- density(data_kde[["intensities"]])#, bw=1)
predict_kde <- function(newdata, data, model) {
  newdata <- sapply(newdata, FUN = function(x) mean(dnorm(x, mean = data, sd = model$bw)))
  return (newdata)
}
newdata_kde <- data.frame("intensities" = seq(discretized$min, discretized$max, length = 2000))
predictions_kde <- data.frame("intensities" = newdata_kde[["intensities"]],
                   "predictions" = predict_kde(newdata = newdata_kde[["intensities"]], 
                                               data = data_kde[["intensities"]], model = model_kde))

ggplot() + 
  geom_histogram(data = intensities, breaks = discretized$breaks,
                 mapping = aes(x = intensities, y = after_stat(density)), 
                 fill = "grey90", color = "black") +
  geom_line(data = predictions_sef, mapping = aes(x = intensities, y = predictions, color = "SEF2")) +
  geom_line(data = predictions_kde, mapping = aes(x = intensities, y = predictions, color = "KDE")) +
  #geom_line(data = predictions_gam, mapping = aes(x = intensities, y = predictions, color = "GAM")) +
  geom_line(data = predictions_sef2, mapping = aes(x = intensities, y = predictions, color = "SEF")) +
  scale_colour_manual(values = c("SEF" = "blue", "KDE" = "red", "SEF2" = "green"), name = "Method") +
  xlab("Intensities") + ylab("Density") + theme_minimal()
```

We create a plot where we fit models of different flexibility on the intensities
of one of the peptides:

```{r}
intensities <- data.frame("intensities" = c(assay(sce[1,])))
discretized <- discretize(intensities, 40)

carrier <- rep(sum(discretized$counts)*(discretized$breaks[2]-discretized$breaks[1]),
               length(discretized$counts))

data <- data.frame("intensities" = discretized$mids,
                  "counts" = discretized$counts, "carrier" = carrier)
newdata <- data.frame("intensities" = seq(discretized$min, discretized$max, length = 2000),
                      "carrier" = rep(1, length = 2000))

model_glm1 <- glm(counts ~ intensities,
             offset = log(carrier), data, family = "poisson")
predictions_glm1 <- data.frame("intensities" = newdata$intensities,
                   "predictions" = predict(model_glm1, newdata = newdata, type = "response"))

model_glm2 <- glm(counts ~ intensities + I(intensities^2),
             offset = log(carrier), data, family = "poisson")
predictions_glm2 <- data.frame("intensities" = newdata$intensities,
                   "predictions" = predict(model_glm2, newdata = newdata, type = "response"))

model_glm3 <- glm(counts ~ intensities + I(intensities^2) + I(intensities^3),
             offset = log(carrier), data, family = "poisson")
predictions_glm3 <- data.frame("intensities" = newdata$intensities,
                   "predictions" = predict(model_glm3, newdata = newdata, type = "response"))

model_glm4 <- glm(counts ~ intensities + I(intensities^2) + I(intensities^3) + I(intensities^4),
             offset = log(carrier), data, family = "poisson")
predictions_glm4 <- data.frame("intensities" = newdata$intensities,
                   "predictions" = predict(model_glm4, newdata = newdata, type = "response"))

model_gam <- gam(counts ~ s(intensities), 
                 offset = log(carrier), data, family = "poisson")
predictions_gam <- data.frame("intensities" = newdata$intensities,
                   "predictions" = predict(model_gam, newdata = newdata, type = "response"))

ggplot() + 
  geom_histogram(data = intensities, breaks = discretized$breaks,
                 mapping = aes(x = intensities, y = after_stat(density)), 
                 fill = "grey90", color = "black") +
  geom_line(data = predictions_glm1, 
            mapping = aes(x = intensities, y = predictions, color = "GLM1")) +
  geom_line(data = predictions_glm2, 
            mapping = aes(x = intensities, y = predictions, color = "GLM2")) +
  geom_line(data = predictions_glm3, 
            mapping = aes(x = intensities, y = predictions, color = "GLM3")) +
  geom_line(data = predictions_glm4, 
            mapping = aes(x = intensities, y = predictions, color = "GLM4")) +
  geom_line(data = predictions_gam, 
            mapping = aes(x = intensities, y = predictions, color = "GAM0")) +
  scale_colour_manual(values = c("GLM1" = "orange", "GLM2" = "green", "GLM3" = "blue", 
                                 "GLM4" = "purple", "GAM0" = "red"), name = "Model") +
  xlab("Intensities") + ylab("Density") + theme_minimal()
```

Now we do the same, but only select the intensities of monocytes:

```{r}
intensities <- data.frame("intensities" = 
                            c(assay(sce[1,][,colData(sce[1,])$SampleType == "Monocyte"])))
discretized <- discretize(intensities, 40)

carrier <- rep(sum(discretized$counts)*(discretized$breaks[2]-discretized$breaks[1]),
               length(discretized$counts))

data <- data.frame("intensities" = discretized$mids,
                  "counts" = discretized$counts, "carrier" = carrier)
newdata <- data.frame("intensities" = seq(discretized$min, discretized$max, length = 2000),
                      "carrier" = rep(1, length = 2000))

model_glm1 <- glm(counts ~ intensities,
             offset = log(carrier), data, family = "poisson")
predictions_glm1 <- data.frame("intensities" = newdata$intensities,
                   "predictions" = predict(model_glm1, newdata = newdata, type = "response"))

model_glm2 <- glm(counts ~ intensities + I(intensities^2),
             offset = log(carrier), data, family = "poisson")
predictions_glm2 <- data.frame("intensities" = newdata$intensities,
                   "predictions" = predict(model_glm2, newdata = newdata, type = "response"))

model_glm3 <- glm(counts ~ intensities + I(intensities^2) + I(intensities^3),
             offset = log(carrier), data, family = "poisson")
predictions_glm3 <- data.frame("intensities" = newdata$intensities,
                   "predictions" = predict(model_glm3, newdata = newdata, type = "response"))

model_glm4 <- glm(counts ~ intensities + I(intensities^2) + I(intensities^3) + I(intensities^4),
             offset = log(carrier), data, family = "poisson")
predictions_glm4 <- data.frame("intensities" = newdata$intensities,
                   "predictions" = predict(model_glm4, newdata = newdata, type = "response"))

model_gam <- gam(counts ~ s(intensities), 
                 offset = log(carrier), data, family = "poisson")
predictions_gam <- data.frame("intensities" = newdata$intensities,
                   "predictions" = predict(model_gam, newdata = newdata, type = "response"))

ggplot() + 
  geom_histogram(data = intensities, breaks = discretized$breaks,
                 mapping = aes(x = intensities, y = after_stat(density)), 
                 fill = "grey90", color = "black") +
  geom_line(data = predictions_glm1, 
            mapping = aes(x = intensities, y = predictions, color = "GLM1")) +
  geom_line(data = predictions_glm2, 
            mapping = aes(x = intensities, y = predictions, color = "GLM2")) +
  geom_line(data = predictions_glm3, 
            mapping = aes(x = intensities, y = predictions, color = "GLM3")) +
  geom_line(data = predictions_glm4, 
            mapping = aes(x = intensities, y = predictions, color = "GLM4")) +
  geom_line(data = predictions_gam, 
            mapping = aes(x = intensities, y = predictions, color = "GAM0")) +
  scale_colour_manual(values = c("GLM1" = "orange", "GLM2" = "green", "GLM3" = "blue", 
                                 "GLM4" = "purple", "GAM0" = "red"), name = "Model") +
  xlab("Intensities") + ylab("Density") + theme_minimal()
```