---
title: "Dimensionality reduction"
author: "Fleur Ballet"
output: html_document
date: "2024-05-27"
---

```{r setup, include = FALSE}
## Options for Rmarkdown compilation
knitr::opts_chunk$set(fig.width = 7,
                      fig.height = 5,
                      fig.align = "center",
                      out.width = "70%",
                      message = FALSE,
                      collapse = TRUE,
                      crop = NULL, warning = FALSE)
```

```{r}
## Single-cell and protoeomics processing tools
library("scpdata")
library("scp")
## Visualization and manipulation tools
library("ggplot2")
library("dplyr")
library("patchwork")
# PCA
library("scater")
```

# Load data

Load the previously processed single-cell experiments.
Select whether the real data (sces) must be used or the mock data (sces_subset_split or sces_subset_split_per_run) by setting the other lines for "path" to comments.

```{r}
loadRData <- function(fileName){
    load(fileName)
    get(ls()[ls() != "fileName"])
}

data_path <- "Data/leduc_sces.Rda"
data_path <- "Data/leduc_sces_subset_split.Rda"
data_path <- "Data/leduc_sces_subset_split_per_run.Rda"

sces <- loadRData(data_path)
```

# PCA

Calculate the principal components for each of the considered assays.
First for the assays of the minimal processing procedure:

```{r}
mxN <- assay(sces[["peptides_log"]])
mxN[is.na(mxN)] <- 0
reducedDim(sces[["peptides_log"]], "PCA") <- runPCA(mxN, rank = 5, scale = TRUE)$rotation

mxN <- assay(sces[["peptides_norm"]])
mxN[is.na(mxN)] <- 0
reducedDim(sces[["peptides_norm"]], "PCA") <- runPCA(mxN, rank = 5, scale = TRUE)$rotation

sces[["peptides_imptd"]] <- runPCA(sces[["peptides_imptd"]], 
                                   ncomponents = 5, ntop = Inf, scale = TRUE,
                                   exprs_values = 1, name = "PCA")

sces[["peptides_batchC"]] <- runPCA(sces[["peptides_batchC"]], 
                                    ncomponents = 5, ntop = Inf, scale = TRUE,
                                    exprs_values = 1, name = "PCA")
```

And for the assays of the alternative summarisation:

```{r}
mxS <- assay(sces[["peptidesSweep"]])
mxS[is.na(mxS)] <- 0             
reducedDim(sces[["peptidesSweep"]], "PCA") <- runPCA(mxS, rank = 5)$rotation
reducedDim(sces[["peptidesSweep"]], "PCA_scale") <- runPCA(mxS, rank = 5, scale = TRUE)$rotation

mxS <- assay(sces[["peptidesSweep"]])
mxS <- sweep(mxS, MARGIN=1, FUN = "-", STATS = rowMedians(mxS, na.rm = TRUE))
mxS <- sweep(mxS, MARGIN=1, FUN = "/", STATS = rowMads(mxS, na.rm = TRUE))
mxS[is.na(mxS)] <- 0             
reducedDim(sces[["peptidesSweep"]], "PCA_scaleMad") <- runPCA(mxS, rank = 5)$rotation

sces[["peptidesSweepImp"]] <- runPCA(sces[["peptidesSweepImp"]], 
                                    ncomponents = 5, ntop = Inf, scale = TRUE,
                                    exprs_values = 1, name = "PCA")
```

# Plots

Now we can plot the PC-projections, where we distinguish the points of different batches and sample types.
First create the plots:

```{r}
plotlist <- list()
for (name in names(sces)) {
  for (dimred in reducedDimNames(sces[[name]])) {
    plotlist[[paste0(paste0(name,": ",dimred))]] <- 
      plotReducedDim(sces[[name]], dimred = dimred, point_alpha = 1,
                     colour_by = "lcbatch", shape_by = "SampleType") +
      ggtitle(paste0(paste0(name,": ",dimred)))
  }
}
```

Show the PC-plots of the different assays in one plot:

```{r}
p1 <- plotlist[["peptides_log: PCA"]]
p2 <- plotlist[["peptides_norm: PCA"]]
p3 <- plotlist[["peptides_imptd: PCA"]]
p4 <- plotlist[["peptides_batchC: PCA"]]
p5 <- plotlist[["peptidesSweep: PCA"]]
p6 <- plotlist[["peptidesSweepImp: PCA"]]
combined <- p1 + p2 + p3 + p4 + p5 + p6 & theme(legend.position = "right")
combined + plot_layout(guides = "collect") #1200x800
```

We make another plot to visualise the set effects, but do this on a reduced dataset.

```{r}
temp<-sces[["peptides_norm"]]
temp<-temp[,colData(temp)[["Set"]]%in%c("eAL00219","wAL00200","eAL00266","wAL00286")]
tempassay<-assay(temp)
tempassay[is.na(tempassay)] <- 0
reducedDim(temp, "PCA") <- runPCA(tempassay, rank = 5, scale = TRUE)$rotation
plotReducedDim(temp, dimred = reducedDimNames(temp), point_alpha = 1,
               colour_by = "Set", shape_by = "SampleType")
```

We now create several PCA plots in one large figure, again for all the assays, where the points are coloured according to different variables (SampleType, lcbatch, Channel and Set).

```{r}
plotlist <- list()
for (name in names(sces)) {
  plot1 <- plotReducedDim(sces[[name]], dimred = "PCA", point_alpha = 1,
                     colour_by = "SampleType", shape_by = "SampleType") + 
    theme(legend.position = 'none')
  plot2 <- plotReducedDim(sces[[name]], dimred = "PCA", point_alpha = 1,
                     colour_by = "lcbatch", shape_by = "SampleType") + 
    theme(legend.position = 'none')
  plot3 <- plotReducedDim(sces[[name]], dimred = "PCA", point_alpha = 1,
                     colour_by = "Channel", shape_by = "SampleType") + 
    theme(legend.position = 'none')
  plot4 <- plotReducedDim(sces[[name]], dimred = "PCA", point_alpha = 1,
                     colour_by = "Set", shape_by = "SampleType") + 
    theme(legend.position = 'none')
  plotlist[[name]] <- wrap_elements(plot1 + plot2 + plot3 + plot4 + plot_layout(ncol = 1) +
    plot_annotation(name, theme = theme(plot.title = 
                                          element_text(hjust = 0.5, face = 'bold', size = 20))))
}
combined <- wrap_plots(plotlist) + plot_layout(ncol = 3, guides = 'collect') + 
    theme(legend.position = 'bottom')  + 
  plot_annotation("PCA of Leduc", theme = theme(plot.title = 
                              element_text(hjust = 0.5, face = 'bold', size = 25))) #1600x2000
```

# Save data

Save the adapted single-cell experiments. Now the principal components are stored in the SingleCellExperiment objects.

```{r}
save(sces, file = data_path)
```

