---
title: "Fitting msqrob2 models"
author: "Fleur Ballet"
output: html_document
date: "2024-05-27"
---

```{r setup, include = FALSE}
## Options for Rmarkdown compilation
knitr::opts_chunk$set(fig.width = 7,
                      fig.height = 5,
                      fig.align = "center",
                      out.width = "70%",
                      message = FALSE,
                      collapse = TRUE,
                      crop = NULL, warning = FALSE)
```

```{r}
library("scpdata")
library("scp")

library("ggplot2")
library("dplyr")
library("patchwork")

library("impute")
library("scater")
library("msqrob2")
library("tibble")
```

Start from the data that was previously preprocessed in "minimal_preprocessing" 
and "preprocessing_continued". Select whether the real data (sces) or the mock data
(sces_subset_split or sces_subset_split_per_run) should be used by putting the other 
lines in comment; also note the right SampleTypeValue should be selected.

```{r}
loadRData <- function(fileName){
    load(fileName)
    get(ls()[ls() != "fileName"])
}

## select for real data
SampleTypeValue <- "SampleTypeMonocyte"
data_path <- "Data/leduc_sces.Rda"
# or
## select for mock data
#SampleTypeValue <- "SampleTypeMonocytetype2"
#data_path <- "Data/leduc_sces_subset_split.Rda" # or
#data_path <- "Data/leduc_sces_subset_split_per_run.Rda"

sces <- loadRData(data_path)
```


# Model protein expression values

Fit the msqrob2 models for each peptide in each assay and perform a hypothesis test.
We fit 3 models. The first (MR1) models the intensity in function of only SampleType:

```{r}
for (name in names(sces)) {
  sces[[name]] <- msqrob(sces[[name]], ~ SampleType, overwrite = TRUE)
  L <- makeContrast(paste0(SampleTypeValue,"=0"), parameterNames = c(SampleTypeValue))
  sces[[name]] <- hypothesisTest(sces[[name]], contrast = L, overwrite = TRUE)
}
```

The second (MR2) adds a random effect for Set:

```{r}
sces2 <- sces
for (name in names(sces2)) {
  sces2[[name]] <- msqrob(sces2[[name]], ~ SampleType + (1|Set), overwrite = TRUE)
  L <- makeContrast(paste0(SampleTypeValue,"=0"), parameterNames = c(SampleTypeValue))
  sces2[[name]] <- hypothesisTest(sces2[[name]], contrast = L, overwrite = TRUE)
}
```

The third (MR3) further includes an effect for Channel:

```{r}
sces3 <- sces
for (name in names(sces3)) {
  sces3[[name]] <- msqrob(sces3[[name]], ~ SampleType + (1|Set), overwrite = TRUE)
  L <- makeContrast(paste0(SampleTypeValue,"=0"), parameterNames = c(SampleTypeValue))
  sces3[[name]] <- hypothesisTest(sces3[[name]], contrast = L, overwrite = TRUE)
}
```

Save the updated assays, in which the models are stored:

```{r}
save(sces, file = data_path)
save(sces2, file = paste0(substr(data_path, 1, nchar(data_path)-4),"2.Rda"))
save(sces3, file = paste0(substr(data_path, 1, nchar(data_path)-4),"3;Rda"))
```

# Results

Look at the results of each model:

```{r}
summaries_msqrob <- list()
for (name in names(sces)) {
  summaries_msqrob[[name]] <- rowData(sces[[name]])[[SampleTypeValue]] %>% 
    arrange(pval) %>% filter(adjPval<0.05)
}
summaries_msqrob
```

Look at the volcanoplots for each assay:

```{r}
volcanoplotlist <- list()
for (name in names(sces)) {
  volcanoplotlist[[name]] <- 
      ggplot(rowData(sces[[name]])[[SampleTypeValue]],
             aes(x = logFC, y = -log10(pval), color = adjPval < 0.05)) +
      geom_point(cex = 2.5) + scale_color_manual(values = alpha(c("black", "red"), 0.5)) +
      theme_minimal() + ggtitle(paste0(name,": msqrob2"))
}
for (plot in volcanoplotlist) {
  print(plot)
}
```

We investigate the distributions of the p-values:

```{r}
histplotlist <- list()
for (name in names(sces)) {
  temp <- as.data.frame(rowData(sces[[name]])[[SampleTypeValue]][["pval"]])
  colnames(temp) <- "pvalue"
  histplotlist[[name]] <- 
    ggplot(data = temp, aes(x = pvalue)) + 
        geom_histogram(color = "black", fill = "gray") + 
        labs(title = name, x = "p-value", y = "Frequency") + theme_minimal()
}
```

We plot the histograms together in one plot:

```{r}
p1 <- histplotlist[["peptides_log"]]
p2 <- histplotlist[["peptides_norm"]]
p3 <- histplotlist[["peptides_imptd"]]
p4 <- histplotlist[["peptides_batchC"]]
p5 <- histplotlist[["peptidesSweep"]]
p6 <- histplotlist[["peptidesSweepImp"]]
combined <- p1 + p2 + p3 + p4 + p5 + p6 & theme(legend.position = "right")
combined + plot_layout(ncol = 2, guides = "collect") # 800x600
```

We can do the same for the distributions of the adjusted p-values:

```{r}
histplotlist <- list()
for (name in names(sces)) {
  temp <- as.data.frame(rowData(sces[[name]])[[SampleTypeValue]][["adjPval"]])
  colnames(temp) <- "pvalue"
  histplotlist[[name]] <- 
    ggplot(data = temp, aes(x = pvalue)) + 
        geom_histogram(color = "black", fill = "gray") + 
        labs(title = name, x = "Adjusted p-value", y = "Frequency") + theme_minimal()
}
```

And again in one plot:

```{r}
p1 <- histplotlist[["peptides_log"]]
p2 <- histplotlist[["peptides_norm"]]
p3 <- histplotlist[["peptides_imptd"]]
p4 <- histplotlist[["peptides_batchC"]]
p5 <- histplotlist[["peptidesSweep"]]
p6 <- histplotlist[["peptidesSweepImp"]]
combined <- p1 + p2 + p3 + p4 + p5 + p6 & theme(legend.position = "right")
combined + plot_layout(ncol = 2, guides = "collect") # 800x600
```

The following function returns the peptides for which a significant difference
between the two sample types was found. The second function prints this in a table.

```{r}
get_significant_peptides <- function(sce, name) {
  sigNames <- rowData(sce)[[SampleTypeValue]] %>%
    rownames_to_column(name) %>%
    filter(adjPval < 0.05) %>%
    pull(name)
  return (sigNames)
}

# To print a nice table with the number of (significant) peptides per assay to export
print_nr_sign_pept <- function(sces) {
  numbers <- list()
  numbers[["Nr significant"]] <- sapply(names(sces), function(name)
                                         length(get_significant_peptides(sces[[name]], name)))
  numbers[["Nr peptides"]] <- sapply(names(sces), function(name)
                                         nrow(rowData(sces[[name]])[[SampleTypeValue]]))
  numbers[["Ratio"]] <- numbers[["Nr significant"]]/numbers[["Nr peptides"]]
  numbers <- as.data.frame(numbers)
  return (numbers)
}
```

Print the number of peptides with a significant sample type effect:

```{r}
nr_sign_pept <- print_nr_sign_pept(sces)
nr_sign_pept
```
