---
title: "Results and model comparisons"
author: "Fleur Ballet"
output: html_document
date: "2024-05-27"
---

```{r setup, include = FALSE}
## Options for Rmarkdown compilation
knitr::opts_chunk$set(fig.width = 7,
                      fig.height = 5,
                      fig.align = "center",
                      out.width = "70%",
                      message = FALSE,
                      collapse = TRUE,
                      crop = NULL, warning = FALSE)
```

In this script we study the peptides for which we found a significant difference in 
expression levels between monocytes and melanoma cells. We compare the peptides that 
were identified as significant by the msqrob2 models and our density estimation method.

Loading the packages:

```{r}
library("ggplot2")
library("dplyr")
library("patchwork")
library("tibble")
library("mgcv")
library("scp")
library("xtable")
```

# Function

We first load all the functions from the previous script ("density estimation - multiple groups"), 
so that we have all the functions to perform the density estimation analysis and 
we can present the results nicely. We also add some functions to print tables with results.

```{r}
loadRData <- function(fileName){
    load(fileName)
    get(ls()[ls() != "fileName"])
}
```

Functions to do the formatting of the data and the fitting of the models:

```{r}
# Split the data into a list of datasets, 
# each containing the peptide intensities for one category of the specified variable
split_intensities_data <- function(sce, data, variable) {
  groups <- unique(colData(sce)[,variable])
  split_data <- list()
  for (group in groups) {
    split_data[[group]] <- data[which(colData(sce)[[variable]] == group)]
  }
  return (split_data)
}

# Discretize the data in n_bars intervals
# Define the intervals based on the full dataset and the counts
# Then count the intensities of the different datasets in split_data
discretize_split <- function(data, split_data, n_bars) {
  # discretize
  min <- min(data)
  max <- max(data)
  bar_width <- (max - min)/n_bars
  breaks <- seq(from = min, to = max, by = bar_width)
  mids <- seq(from = min+bar_width/2, length.out = length(breaks)-1, by = bar_width)
  groups <- names(split_data)
  
  # count the intensities in each interval for each category
  counts <- list()
  for (group in groups) {
    counts[[group]] <- sapply(1:(length(breaks)-1), 
                   function(b) sum(split_data[[group]] > breaks[b] & 
                                     split_data[[group]] < breaks[b+1]))
  }
  
  # store results
  results <- list(n_bars, min, max, bar_width, breaks, mids, counts, groups)
  names(results) <- c("n_bars", "min", "max", "bar_width", "breaks", "mids", "counts", "groups")
  return (results)
}

# Format the data so that it can be passed to fit the models
# The models need the sample type, intensities and carrier
format_data_split_long <- function(discretized, carrier = "uniform") {
  length <- length(discretized$mids)
  data_long <- list()
  for (group in discretized$groups) {
    subdata <- list()
    subdata[["SampleType"]] <- rep(group, length)
    subdata[["intensities"]] <- discretized$mids
    subdata[["counts"]] <- discretized$counts[[group]]
    if (carrier == "uniform") {
      subdata[["carrier"]] <- rep(sum(discretized$counts[[group]])*
                                      (discretized$breaks[2]-discretized$breaks[1]), length)
    }
    data_long <- rbind(data_long, as.data.frame(subdata))
  }
  data_long <- as.data.frame(data_long)
  return (data_long)
}

# Fit a Poisson model
# The data need to be formatted as after applying format_data_split_long
# Possible model types:
# - Default is "glm": pass the desired predictors
# - If "gam": the predictors will be smoothed; if "by" variable is passed, this is included
fit_model <- function(data, predictors, type = "glm", by = NULL) {
  if (type == "glm") {
    model <- glm(as.formula(paste("counts", paste(predictors, collapse = "+"), sep = "~")), 
                 data, offset = log(carrier), family = "poisson")
  } else if (type == "gam") {
    model <- gam(as.formula(paste0("counts ~ s(", paste(predictors, collapse = ","), 
                                   paste(", by =", paste0(by, ")")))), 
                 data, offset = log(carrier), family = "poisson")
  }
  return (model)
}

# Make and format the new prediction data and immediately add the made predictions
# "length" specifies the number of new data points in the interval of "discretized"
# Alse pass the model with which we want to make predictions
format_prediction_data_split_long <- function(discretized, length, model, carrier = "uniform") {
  # define the new data
  data_long <- list()
  for (group in discretized$groups) {
    subdata <- list()
    subdata[["SampleType"]] <- rep(group, length)
    subdata[["intensities"]] <- seq(discretized$min, discretized$max, length = length)
    if (carrier == "uniform") {
      subdata[["carrier"]] <- rep(1, length)
    }
    data_long <- rbind(data_long, as.data.frame(subdata))
  }
  data_long <- as.data.frame(data_long)
  
  # do the predictions
  data_long[["predictions"]] <- predict(model, newdata = data_long, type = "response")
  return (data_long)
}

# Transform the split (as with split_intensities_data) intensity-assays to long data
format_intensities_long <- function(split_intensities) {
  data_long <- list()
  for (group in names(split_intensities)) {
    length <- length(split_intensities[[group]])
    subdata <- list()
    subdata[["SampleType"]] <- rep(group, length)
    subdata[["intensities"]] <- split_intensities[[group]]
    data_long <- rbind(data_long, as.data.frame(subdata))
  }
  data_long <- as.data.frame(data_long)
  return (data_long)
}

# Fit multiple models
# "data" need to be the data as formatted with format_data_split_long
# "degrees" is the order until which the main effects for "intensities" need to be included in the model
# "degrees_interactions" is the order until which the interactions of intensities with SampleType need to be included
# if "smooth": also a gam model will be fitted
# Return a list of the models
fit_models_all_degrees <- function(data, degrees, degrees_interactions, smooth) {
  models <- list()
  for (degree in degrees) {
    predictors <- c("SampleType")
    if (degree >= 1) {
      predictors <- append(predictors, "intensities")
    }
    if (degree > 1) {
      for (deg in 2:degree) {
        predictors <- append(predictors, paste0(paste0("I(intensities^",deg),")"))
      }
    }
    if (length(degrees_interactions) > 0) {
      if (1 %in% degrees_interactions) {
        models[[paste0("GLM",deg,0)]] <- fit_model(data, predictors)
      }
      predictors <- append(predictors, "SampleType:intensities")
      if (1 %in% degrees_interactions) {
        models[[paste0("GLM",deg,1)]] <- fit_model(data, predictors)
      }
    }
    for (deg_inter in degrees_interactions) {
      if (deg_inter > 1) {
        predictors <- append(predictors, paste0(paste0("SampleType:I(intensities^",deg_inter),")"))
        models[[paste0("GLM",deg,deg_inter)]] <- fit_model(data, predictors)
      }
    }
  }
  if (smooth) {
    models[["GAM"]] <- fit_model(data, c("intensities"), "gam", "as.factor(SampleType)")
    models[["GAM0"]] <- fit_model(data, c("intensities"), "gam")
  }
  return (models)
}
```

Functions for hypothesis testing:

```{r}
# Calculate the deviance of the models
get_deviance_models <- function(models) {
  print("Deviance:")
  for (name in names(models)) {
    cat(name,": ",deviance(models[[name]]),"\n")
  }
} 

# Perform a deviance test for the model
deviance_test <- function(model) {
  return (pchisq(model$deviance, model$df.residual, lower.tail = FALSE))
}

# Perform a Wald test for the model, 
# where we test for the combination of parameters in "coeffs"
# This function is not applicable for the gam models
wald_test <- function(model, coeffs) {
  indices <- which(names(coef(model)) %in% coeffs)
  contrast <- cbind(matrix(rep(0, length(indices)*length(coef(model))), nrow = length(indices)))
  for (i in 1:length(indices)) {
    contrast[i,indices[i]] <- 1
  }
  test_stat <- t(contrast %*% coef(model)) %*% 
    solve(contrast %*% vcov(model) %*% t(contrast)) %*% 
    (contrast %*% coef(model))
  p_value <- pchisq(test_stat, df = length(indices), lower.tail = FALSE)
  return (p_value)
}

# Perform a likelihood ratio test for a model compared to a null model
lr_test <- function(model, null_model) {
  test <- anova(null_model, model, test = "Chisq")
  p_value <- test[2, "Pr(>Chi)"]
  return (p_value)
}

# Perform a deviance, likelihood ratio or Wald test for all the models
# return the p-values
test_all <- function(models, test) {
  p_values <- list()
  for (name in names(models)[sapply(names(models), function(x) !(x %in% c("GLM40","GAM0")))]) {
    model <- models[[name]]
    if (test == "deviance_test") {
      p_values[[name]] <- match.fun(test)(model)
    } else if (test == "lr_test") {
      if (name == "GAM") {
        null_model <- models[["GAM0"]]
      } else {
        null_model <- models[["GLM40"]]
      }
      p_values[[name]] <- match.fun(test)(model, null_model)
    } else if (test == "wald_test") {
      if (name == "GAM") {
        p_values[[name]] <- NA
      } else{
        coeffs <- names(coef(model))[unlist(lapply(names(coef(model)), function(x) grepl(":", x)))]
        p_values[[name]] <- match.fun(test)(model, coeffs)
      }
    }
  }
  p_values <- cbind(p_values, p.adjust(p_values, method = "BH"))
  colnames(p_values) <- c("P-value", "Adj-P-value")
  return (p_values)
}

# Print a list nicely
print_list <- function(list, name) {
  print(name)
  for (item in names(list)) {
    cat(item,": ",list[[item]],"\n")
  }
}

# Print the p-values of the different tests together to get a nice table to export
print_all_pvalues <- function(list_p) {
  p_values <- do.call(cbind, list_p)
  colnames(p_values) <- sapply(1:length(colnames(p_values)), 
                               function(i) paste(names(list_p)[(i+1)/2], colnames(p_values)[i]))
  return (p_values)
}

# Print the p-values of the different tests together to get a nice table to export,
# but drop the adjusted p-values column
print_all_pvalues_only <- function(list_p) {
  p_values <- do.call(cbind, list_p)[,c(1,3,5)]
  colnames(p_values) <- sapply(1:length(colnames(p_values)), 
                               function(i) paste(names(list_p)[(i+1)/2], colnames(p_values)[i]))
  return (p_values)
}

# Print the p-values of the msqrob2 models together
print_pvalues_msqrob <- function(list_p, peptide) {
  p_values <- rbind("MR1" = list_p[["MR1"]][peptide,],
        "MR2" = list_p[["MR2"]][peptide,], 
        "MR3" = list_p[["MR3"]][peptide,])
  return (p_values)
}
```

Function to plot the predictions of the models on the original histograms:

```{r}
# Plot the histograms for all the categories of sample type
# Plot the predictions for the given model for all categories on top
# The data need to be formatted as with split_intensities_long
plot_intensities_prediction <- function(split_intensities_long, discretized, model, name = NULL) {
  newdata_long <- format_prediction_data_split_long(discretized, 2000, model)
  
  ggplot() + 
    geom_histogram(data = split_intensities_long, breaks = discretized$breaks,
                   mapping = aes(x = intensities, y = after_stat(density), 
                             fill = SampleType, color = SampleType, alpha = SampleType),
                   position = "dodge") + 
    geom_line(data = newdata_long, aes(x = intensities, y = predictions, color = SampleType)) +
    scale_colour_manual(values = c("darkblue", "red")) +
    scale_fill_manual(values = c("lightblue", "pink")) +
    scale_alpha_manual(values = c(0.5, 0.5)) +
    xlab("Intensities") + ylab("Density") +
    ggtitle(name) + theme_minimal()
}

# Plot the histograms and prediction for all the models
# returns a list with the plots
plot_all_models <- function(split_intensities_long, discretized, models) {
  plotlist <- list()
  for (name in names(models)[sapply(names(models), function(x) !(x %in% c("GLM40","GAM0")))]) {
    plotlist[[name]] <- plot_intensities_prediction(split_intensities_long, 
                                                    discretized, models[[name]], name)
  }
  return (plotlist)
}
```

Functions to apply the full procedure to fit the models. We can immediately test or plot the fits.
These can be applied to a desired assay and peptide.

```{r}
# Fit Poisson models on the intensities of a certain peptide in a certain assay sce
# It fits GLM models where the main effect of "intensities" is included up to the 4th order
# and also "intensities" to the 4th order in interactions with SampleType
# It also fit GAM models for each SampleType
# Returns a list with the p-values
model_and_test <- function(peptide, sce) {
  
  # Prepare the data
  intensities <- assay(sce[peptide,])
  split_intensities <- split_intensities_data(sce, intensities, "SampleType")
  split_intensities_long <- format_intensities_long(split_intensities)
  discretized <- discretize_split(intensities, split_intensities, 40)
  data_long <- format_data_split_long(discretized)
  
  # Fit the models
  models <- fit_models_all_degrees(data_long, 4, 0:4, TRUE)
  
  # Test significance SampleType
  p_dev <- test_all(models, "deviance_test")
  p_wald <- test_all(models, "wald_test")
  p_lr <- test_all(models, "lr_test")
  
  results <- list("p_dev" = p_dev, "p_wald" = p_wald, "p_lr" = p_lr)
  return (results)
}

# Fit Poisson models on the intensities of all the peptides in a certain assay sce
# It fits GLM models where the main effect of "intensities" is included up to the 4th order
# and also "intensities" to the 4th order in interactions with SampleType
# It also fit GAM models for each SampleType
# Returns a list with the plots
model_and_test_all <- function(sce) {
  results <- list()
  for (peptide in rownames(sce)) {
    results[[peptide]] <- model_and_test(peptide, sce)
  }
  return (results)
}

# Extract the p-values from the tests on the models 
get_p_values <- function(results, type) {
  if (type == "dev") {
    p_values <- lapply(results, function(x) x$p_dev)
  } else if (type == "wald") {
    p_values <- lapply(results, function(x) x$p_wald)
  } else if (type == "lr") {
    p_values <- lapply(results, function(x) x$p_lr)
  }
  names(p_values) <- names(results)
  return (p_values)
}


# Extract the p-values of a certain model over the different peptides
get_p_values_per_model <- function(results, type) {
  p_values_all <- get_p_values(results, type)
  p_values <- list()
  for (name in rownames(results[[1]][[1]])) {
    p_values_model <- do.call(rbind, sapply(p_values_all, 
                                                         function(x) x[name, "P-value"]))
    p_values_model <- cbind(p_values_model, p.adjust(p_values_model, method = "BH"))
    colnames(p_values_model) <- c("P-value", "Adj-P-value")
    p_values[[name]] <- p_values_model
  }
  return (p_values)
}

# Extract the peptides with significant p-values per model
get_significant_peptides_per_model <- function(p_values, model) {
  sigNames <- p_values[[model]] %>% as.data.frame %>% rownames_to_column("peptides") %>%
    filter(`Adj-P-value` < 0.05) %>% pull("peptides")
  return (sigNames)
}

# Extract the peptides with significant p-values of all models
get_significant_peptides <- function(p_values) {
  sigNames <- list()
  sigNames <- sapply(names(p_values), function(name)
                                       get_significant_peptides_per_model(p_values, name))
  return (sigNames)
}

# Extract the peptides without significant p-values per model
get_unsignificant_peptides_per_model <- function(p_values, model) {
  sigNames <- p_values[[model]] %>% as.data.frame %>% rownames_to_column("peptides") %>%
    filter(`Adj-P-value` >= 0.05) %>% pull("peptides")
  return (sigNames)
}

# Extract the peptides without significant p-values of all models
get_unsignificant_peptides <- function(p_values) {
  sigNames <- list()
  sigNames <- sapply(names(p_values), function(name)
                                       get_unsignificant_peptides_per_model(p_values, name))
  return (sigNames)
}

# To print a nice table with the number of (significant) peptides per assay to export
print_nr_sign_pept <- function(p_values) {
  numbers <- list()
  numbers[["Nr significant"]] <- sapply(names(p_values), function(name)
                                  length(get_significant_peptides_per_model(p_values, name)))
  numbers[["Nr peptides"]] <- sapply(names(p_values), function(name)
                                         nrow(p_values[[name]]))
  numbers[["Ratio"]] <- numbers[["Nr significant"]]/numbers[["Nr peptides"]]
  numbers <- as.data.frame(numbers)
  return (numbers)
}
```

Functions for plotting the intensities: 

```{r}
# Fit a Poisson model on the intensities of a certain peptide in a certain assay sce
# Model types can be "glm" and "gam"
# Add the desired predictors
fit_and_plot_intensities <- function(peptide, sce, predictors, type = "glm", by = NULL) {
  
  # Prepare the data
  intensities <- assay(sce[peptide,])
  split_intensities <- split_intensities_data(sce, intensities, "SampleType")
  split_intensities_long <- format_intensities_long(split_intensities)
  discretized <- discretize_split(intensities, split_intensities, 40)
  data_long <- format_data_split_long(discretized)
  
  # Fit the model
  model <- fit_model(data = data_long, predictors, type, by)
  
  # Make the predictions
  plot_intensities_prediction(split_intensities_long, discretized, model)
}

# Fit Poisson models on the intensities of a certain peptide in a certain assay sce
# It fits GLM models where the main effect of "intensities" is included up to the 4th order
# and also "intensities" to the 4th order in interactions with SampleType
# It also fit GAM models for each SampleType
# Returns a list with the plots
fit_all_and_plot_intensities <- function(peptide, sce) {
  
  # Prepare the data
  intensities <- assay(sce[peptide,])
  split_intensities <- split_intensities_data(sce, intensities, "SampleType")
  split_intensities_long <- format_intensities_long(split_intensities)
  discretized <- discretize_split(intensities, split_intensities, 40)
  data_long <- format_data_split_long(discretized)
  
  # Fit the models
  models <- fit_models_all_degrees(data_long, 4, 0:4, TRUE)
  
  plotlist <- list()
  for (name in names(models)[sapply(names(models), function(x) !(x %in% c("GLM40","GAM0")))]) {
    plotlist[[name]] <- plot_intensities_prediction(split_intensities_long, 
                                                    discretized, models[[name]], name)
  }
  return (plotlist)
}
```

Functions for plotting the p-values:

```{r}
# Make a histogram of the p-values of a certain model
plot_p_values <- function(p_values, name, type = "P-value", test = "") {
  d <- as.data.frame(p_values[[name]][,type])
  colnames(d) <- c("p_values")
  ggplot(data = d, aes(x = p_values)) + 
        geom_histogram(color = "black", fill = "gray") + 
        labs(title = paste0(test," test: ",name), x = type, y = "Frequency") + theme_minimal()
}

# Make for every model a histogram of the p-values
plot_p_values_all <- function(p_values, type = "P-value", test = "") {
  plotlist <- list()
  for (name in names(p_values)) {
    plotlist[[name]] <- plot_p_values(p_values, name, type, test)
  }
  return (plotlist)
}
```


# Application

## Load and prepare data

We load the data and the results from the msqrob2 and density estimation methods. 
The msqrob2 results are generated in the script "msqrob2 models".
The density results are generated in "density estimation - multiple groups".
The msqrob2 results of the three models are stored in the SingleCellExperiment itself, 
the p-values of the tests on the densities are stored separately.
Select which data needs to be used: the real data (sces) or one of the mock datasets
(sces_subset_split or sces_subset_split_per_run). The same assay as for the density
estimation must be used (here peptidesSweepImp).

```{r}
## Select which data must be used
data_type <- "sces" # real data
#data_type <- "sces_subset_split" # mock data - random fake sample types to each entry
#data_type <- "sces_subset_split_per_run" # mock data - random fake sample types to whole runs
assay <- "peptidesSweepImp"

# Load msqrob2 results
sces <- loadRData(paste0("Data/leduc_",data_type,".Rda")) # for MR1
sce <- sces[[assay]]
sces2 <- loadRData(paste0("Data/leduc_",data_type,"2.Rda")) # for MR2
sce2 <- sces2[[assay]]
sces3 <- loadRData(paste0("Data/leduc_",data_type,"3.Rda")) # for MR3
sce3 <- sces3[[assay]]
#remove(sces2,sces3) # not longer necessary

# Load density results
results_density <- loadRData(paste0("Results/leduc_",data_type,"_results_sef_all.Rda"))

# Small adaptions for compatibility
if (data_type == "sces") {
  colData(sce)[["SampleType"]] <- ifelse(colData(sce)[["SampleType"]] == "Melanoma cell", 
                                         "Melanoma", "Monocyte") # rename for easier use
  SampleTypeValue <- "SampleTypeMonocyte"
} else if (data_type == "sces_subset_split" | data_type == "sces_subset_split_per_run") {
  SampleTypeValue <- "SampleTypeMonocytetype2"
}
```

We reformat the results of the density estimation and the msqrob2 modelling, 
so that they are in the same format and can be used for the same functions.

```{r}
# Density
p_values_dev <- get_p_values_per_model(results_density, "dev")
p_values_wald <- get_p_values_per_model(results_density, "wald")
p_values_lr <- get_p_values_per_model(results_density, "lr")
p_values_density <- list("Deviance" = p_values_dev, "Wald" = p_values_wald, "LR" = p_values_lr)

# Msqrob2
p_values_mr1 <- rowData(sce)[[SampleTypeValue]][, c("pval", "adjPval")]
colnames(p_values_mr1) <- c("P-value", "Adj-P-value")
p_values_mr2 <- rowData(sce2)[[SampleTypeValue]][, c("pval", "adjPval")]
colnames(p_values_mr2) <- c("P-value", "Adj-P-value")
p_values_mr3 <- rowData(sce3)[[SampleTypeValue]][, c("pval", "adjPval")]
colnames(p_values_mr3) <- c("P-value", "Adj-P-value")
p_values_msqrob <- list("MR1" = p_values_mr1, "MR2" = p_values_mr2, "MR3" = p_values_mr3)

# All
p_values_all <- c(p_values_density, "msqrob2" = list(p_values_msqrob))
```

Now p_values_all is a list with as first element the p-values of the deviance test (p_values_dev),
as second those of the Wald test (p_values_wald), as third those of the LR-test (p_values_lr)
and lastly those of the msqrob2 (p_values_msqrob). 
The object of p-values per test (e.g.\ p_values_dev) is a list of the p-values per model.
For the tests on the density models the elements are the p-values for all peptides
for respectively GLM41, GLM42, GLM43, GLM44 and GAM. For p_values_msqrob
the elements are the p-values for MR1, MR2 and MR3.

## Number of significant test results

Get the peptides for which each test gave a significant result.
For the deviance test this are the peptides for which the model had a lack of fit.
For the other tests this are the peptides with a significant SampleType effect 
according to the models. Then print the numbers.

```{r}
sign_dev <- get_significant_peptides(p_values_dev)
sign_wald <- get_significant_peptides(p_values_wald)
sign_lr <- get_significant_peptides(p_values_lr)
sign_msqrob <- get_significant_peptides(p_values_msqrob)
sign_all <- list("Deviance" = sign_dev, "Wald" = sign_wald, "LR" = sign_lr, "msqrob2" = sign_msqrob)

print_nr_sign_pept(p_values_dev)
print_nr_sign_pept(p_values_wald)
print_nr_sign_pept(p_values_lr)
print_nr_sign_pept(p_values_msqrob)
```

Order the peptides according to their p-value in the test for the SampleType effect.
This will be useful later.

```{r}
ordered_glm44 <- p_values_lr[["GLM44"]][,"P-value"][order(
                              unlist(p_values_lr[["GLM44"]][,"P-value"]), decreasing = F)]
ordered_gam <- p_values_lr[["GAM"]][,"P-value"][order(
                              unlist(p_values_lr[["GAM"]][,"P-value"]), decreasing = F)]

temp <- p_values_msqrob[["MR3"]][,"P-value"]
names(temp) <- rownames(p_values_msqrob[["MR3"]])
ordered_mr3 <- temp[order(unlist(temp), decreasing = F)]
```

## Comparison of the results between models - compare numbers

We compare the peptides that were significant according to the LR test on the densities but not according to the msqrob2 models and vice versa. 
Select the density and msqrob model that should be compared; here we choose GLM44 and MR3.

```{r}
model_dens <- "GLM44"
model_msqrob <- "MR3"

common <- sign_lr[[model_dens]][sapply(sign_lr[[model_dens]], function(x) x %in% sign_msqrob[[model_msqrob]]) == TRUE]
print(paste0("Peptides selected by both: ", length(common)))

only_dens <- sign_lr[[model_dens]][sapply(sign_lr[[model_dens]], 
                                          function(x) !(x %in% sign_msqrob[[model_msqrob]])) == TRUE]
print(paste0("Peptides selected only by density: ", length(only_dens)))
only_msqrob <- sign_msqrob[[model_msqrob]][sapply(sign_msqrob[[model_msqrob]], 
                                                function(x) !(x %in% sign_lr[[model_dens]])) == TRUE]
print(paste0("Peptides selected only by msqrob: ", length(only_msqrob)))

neither <- rownames(p_values_lr[[model_dens]])[sapply(rownames(p_values_lr[[model_dens]]), function(x)
  !(x %in% sign_lr[[model_dens]]) & !(x %in% sign_msqrob[[model_msqrob]])) == TRUE]
print(paste0("Peptides not selected: ", length(neither)))
```

We do the same for another density model; we compare GAM and MR3.

```{r}
model_dens_alt <- "GAM"
model_msqrob_alt <- "MR3"

common_alt <- sign_lr[[model_dens_alt]][sapply(sign_lr[[model_dens_alt]], function(x) x %in% sign_msqrob[[model_msqrob_alt]]) == TRUE]
print(paste0("Peptides selected by both: ", length(common_alt)))

only_dens_alt <- sign_lr[[model_dens_alt]][sapply(sign_lr[[model_dens_alt]], 
                                      function(x) !(x %in% sign_msqrob[[model_msqrob_alt]])) == TRUE]
print(paste0("Peptides selected only by density: ", length(only_dens_alt)))
only_msqrob_alt <- sign_msqrob[[model_msqrob_alt]][sapply(sign_msqrob[[model_msqrob_alt]], 
                                        function(x) !(x %in% sign_lr[[model_dens_alt]])) == TRUE]
print(paste0("Peptides selected only by msqrob: ", length(only_msqrob_alt)))

neither_alt <- rownames(p_values_lr[[model_dens_alt]])[sapply(rownames(p_values_lr[[model_dens_alt]]), function(x)
  !(x %in% sign_lr[[model_dens_alt]]) & !(x %in% sign_msqrob[[model_msqrob_alt]])) == TRUE]
print(paste0("Peptides not selected: ", length(neither_alt)))
```

Similarly we could also compare the peptides that were significant according to 
the LR test on one density model and not on the other and vice versa.

```{r}
model_dens1 <- "GLM44"
model_dens2 <- "GAM"

common_dens <- sign_lr[[model_dens1]][sapply(sign_lr[[model_dens1]], function(x) x %in% sign_lr[[model_dens2]]) == TRUE]
print(paste0("Peptides selected by both: ", length(common_dens)))

only_dens1 <- sign_lr[[model_dens1]][sapply(sign_lr[[model_dens1]], 
                                          function(x) !(x %in% sign_lr[[model_dens2]])) == TRUE]
print(paste0("Peptides selected only by density 1: ", length(only_dens1)))
only_dens2 <- sign_lr[[model_dens2]][sapply(sign_lr[[model_dens2]], 
                                                function(x) !(x %in% sign_lr[[model_dens1]])) == TRUE]
print(paste0("Peptides selected only by density 2: ", length(only_dens2)))

neither_dens <- rownames(p_values_lr[[model_dens1]])[sapply(rownames(p_values_lr[[model_dens1]]), function(x)
  !(x %in% sign_lr[[model_dens1]]) & !(x %in% sign_lr[[model_dens2]])) == TRUE]
print(paste0("Peptides not selected: ", length(neither_dens)))
```

## Comparison of the results between models - compare plots and p-values

Study some peptides that were only found significant by the density estimation method. 
Study the most, middle and least significant peptides.

We here look at their p-values in every model:

```{r}
pval <- sort(ordered_glm44[only_dens])
most_sign <- pval[1:3]
middle_sign <- pval[(floor(length(pval)/2)-1):(floor(length(pval)/2)+1)]
least_sign <- pval[(length(pval)-2):length(pval)]
pepts <- c(most_sign, middle_sign, least_sign)

table <- data.frame()
for (peptide in names(pepts)) {
  table[peptide, "GLM44 P"] <- p_values_lr[["GLM44"]][peptide,"P-value"]
  table[peptide, "GLM44 AdjP"] <- p_values_lr[["GLM44"]][peptide,"Adj-P-value"]
  table[peptide, "GAM P"] <- p_values_lr[["GAM"]][peptide,"P-value"]
  table[peptide, "GAM AdjP"] <- p_values_lr[["GAM"]][peptide,"Adj-P-value"]
  table[peptide, "MR3 P"] <- p_values_msqrob[["MR3"]][peptide,"P-value"]
  table[peptide, "MR3 AdjP"] <- p_values_msqrob[["MR3"]][peptide,"Adj-P-value"]
}
table
table%>%xtable(digits = -3)
```

We here look at the histograms of their intensities, and the corresponding densities
estimated by either GAM or GLM44 (choose which by uncommenting the lines).

```{r}
plotlist <- list()
for (peptide in names(pepts)) {
  #plotlist[[peptide]] <- fit_and_plot_intensities(peptide, sce, c("intensities"), "gam",
  #                                              "as.factor(SampleType)") + ggtitle(peptide)
  predictors <- c("SampleType", "intensities", "I(intensities^2)", "I(intensities^3)", 
     "I(intensities^4)", "SampleType:intensities", "SampleType:I(intensities^2)",
     "SampleType:I(intensities^3)", "SampleType:I(intensities^4)")
  plotlist[[peptide]] <- fit_and_plot_intensities(peptide, sce, predictors) + ggtitle(peptide)
}

combined <- wrap_plots(plotlist) & theme(legend.position = "bottom")
combined + plot_layout(ncol = 3, nrow = 3, guides = "collect")
```

Study some peptides that were only found significant by the msqrob2 method.
Like before, we look at the most, middle and least significant peptides.
We first study their p-values:

```{r}
pval <- sort(ordered_mr3[only_msqrob])
most_sign <- pval[1:3]
middle_sign <- pval[(floor(length(pval)/2)-1):(floor(length(pval)/2)+1)]
least_sign <- pval[(length(pval)-2):length(pval)]
pepts <- c(most_sign, middle_sign, least_sign)

table <- data.frame()
for (peptide in names(pepts)) {
  table[peptide, "GLM44 P"] <- p_values_lr[["GLM44"]][peptide,"P-value"]
  table[peptide, "GLM44 AdjP"] <- p_values_lr[["GLM44"]][peptide,"Adj-P-value"]
  table[peptide, "GAM P"] <- p_values_lr[["GAM"]][peptide,"P-value"]
  table[peptide, "GAM AdjP"] <- p_values_lr[["GAM"]][peptide,"Adj-P-value"]
  table[peptide, "MR3 P"] <- p_values_msqrob[["MR3"]][peptide,"P-value"]
  table[peptide, "MR3 AdjP"] <- p_values_msqrob[["MR3"]][peptide,"Adj-P-value"]
}
table
table%>%xtable(digits = -3)

table <- data.frame()
for (peptide in names(pepts)) {
  table[peptide, "MR3 FC"] <- rowData(sce3)[[SampleTypeValue]][peptide,"logFC"]
}
table
table%>%xtable(digits = -3)
```

And then their estimated densities (again fitted with GAM or GLM44):

```{r}
plotlist <- list()
for (peptide in names(pepts)) {
  #plotlist[[peptide]] <- fit_and_plot_intensities(peptide, sce, c("intensities"), "gam",
  #                                              "as.factor(SampleType)") + ggtitle(peptide)
  predictors <- c("SampleType", "intensities", "I(intensities^2)", "I(intensities^3)", 
     "I(intensities^4)", "SampleType:intensities", "SampleType:I(intensities^2)",
     "SampleType:I(intensities^3)", "SampleType:I(intensities^4)")
  plotlist[[peptide]] <- fit_and_plot_intensities(peptide, sce, predictors) + ggtitle(peptide)
}

combined <- wrap_plots(plotlist) & theme(legend.position = "bottom")
combined + plot_layout(ncol = 3, nrow = 3, guides = "collect")
```

Study the peptides for which there was no significant result for both models.
First their p-values:

```{r}
pval <- sort(ordered_gam[neither])
most_sign <- pval[1:3]
middle_sign <- pval[(floor(length(pval)/2)-1):(floor(length(pval)/2)+1)]
least_sign <- pval[(length(pval)-2):length(pval)]
pepts <- c(most_sign, middle_sign, least_sign)

table <- data.frame()
for (peptide in names(pepts)) {
  table[peptide, "GLM44 P"] <- p_values_lr[["GLM44"]][peptide,"P-value"]
  table[peptide, "GLM44 AdjP"] <- p_values_lr[["GLM44"]][peptide,"Adj-P-value"]
  table[peptide, "GAM P"] <- p_values_lr[["GAM"]][peptide,"P-value"]
  table[peptide, "GAM AdjP"] <- p_values_lr[["GAM"]][peptide,"Adj-P-value"]
  table[peptide, "MR3 P"] <- p_values_msqrob[["MR3"]][peptide,"P-value"]
  table[peptide, "MR3 AdjP"] <- p_values_msqrob[["MR3"]][peptide,"Adj-P-value"]
}
table
table%>%xtable(digits = -3)
```

And their densities:

```{r}
plotlist <- list()
for (peptide in names(pepts)) {
  plotlist[[peptide]] <- fit_and_plot_intensities(peptide, sce, c("intensities"), "gam",
                                                "as.factor(SampleType)") + ggtitle(peptide)
  predictors <- c("SampleType", "intensities", "I(intensities^2)", "I(intensities^3)", 
     "I(intensities^4)", "SampleType:intensities", "SampleType:I(intensities^2)",
     "SampleType:I(intensities^3)", "SampleType:I(intensities^4)")
  #plotlist[[peptide]] <- fit_and_plot_intensities(peptide, sce, predictors) + ggtitle(peptide)
}

combined <- wrap_plots(plotlist) & theme(legend.position = "bottom")
combined + plot_layout(ncol = 3, nrow = 3, guides = "collect")
```

Study the peptides with the most significant result (according to GLM44 or GAM).
First look at the p-values for the effect of sample type:

```{r}
#pval <- sort(ordered_glm44[sign_lr[["GLM44"]]])
pval <- sort(ordered_gam[sign_lr[["GAM"]]])
most_sign <- pval[1:3]
middle_sign <- pval[(floor(length(pval)/2)-1):(floor(length(pval)/2)+1)]
least_sign <- pval[(length(pval)-2):length(pval)]
pepts <- c(most_sign, middle_sign, least_sign)

table <- data.frame()
for (peptide in names(pepts)) {
  table[peptide, "GLM44 P"] <- p_values_lr[["GLM44"]][peptide,"P-value"]
  table[peptide, "GLM44 AdjP"] <- p_values_lr[["GLM44"]][peptide,"Adj-P-value"]
  table[peptide, "GAM P"] <- p_values_lr[["GAM"]][peptide,"P-value"]
  table[peptide, "GAM AdjP"] <- p_values_lr[["GAM"]][peptide,"Adj-P-value"]
  table[peptide, "MR3 P"] <- p_values_msqrob[["MR3"]][peptide,"P-value"]
  table[peptide, "MR3 AdjP"] <- p_values_msqrob[["MR3"]][peptide,"Adj-P-value"]
}
table
table%>%xtable(digits = -3)
```

Check whether the density models had a lack of fit or not:

```{r}
table <- data.frame()
for (peptide in names(pepts)) {
  table[peptide, "GLM44 P"] <- p_values_dev[["GLM44"]][peptide,"P-value"]
  table[peptide, "GLM44 AdjP"] <- p_values_dev[["GLM44"]][peptide,"Adj-P-value"]
  table[peptide, "GAM P"] <- p_values_dev[["GAM"]][peptide,"P-value"]
  table[peptide, "GAM AdjP"] <- p_values_dev[["GAM"]][peptide,"Adj-P-value"]
}
table
table%>%xtable(digits = -3)
```

Now study their densities (fitted either with GAM or GLM44).

```{r}
plotlist <- list()
for (peptide in names(pepts)) {
  #plotlist[[peptide]] <- fit_and_plot_intensities(peptide, sce, c("intensities"), "gam",
  #                                              "as.factor(SampleType)") + ggtitle(peptide)
  predictors <- c("SampleType", "intensities", "I(intensities^2)", "I(intensities^3)", 
     "I(intensities^4)", "SampleType:intensities", "SampleType:I(intensities^2)",
     "SampleType:I(intensities^3)", "SampleType:I(intensities^4)")
  plotlist[[peptide]] <- fit_and_plot_intensities(peptide, sce, predictors) + ggtitle(peptide)
}

combined <- wrap_plots(plotlist) & theme(legend.position = "bottom")
combined + plot_layout(ncol = 3, nrow = 3, guides = "collect")
```

Look at a peptide with a very significant result:

```{r}
p_values_dev[["GAM"]]["SDEAVKPFGLK",]
p_values_lr[["GAM"]]["SDEAVKPFGLK",]
p_values_dev[["GLM44"]]["SDEAVKPFGLK",]
p_values_lr[["GLM44"]]["SDEAVKPFGLK",]

print_pvalues_msqrob(p_values_msqrob, "SDEAVKPFGLK")

plotlist <- fit_all_and_plot_intensities("SDEAVKPFGLK", sce)
combined <- wrap_plots(plotlist) & theme(legend.position = "right")
combined + plot_layout(ncol = 2, guides = "collect") # 1200x700 or 1200x1000
```

Study the peptides for which there was no significant effect according to GAM 
in the density estimation:

```{r}
unsign_gam <- get_unsignificant_peptides_per_model(p_values_lr, "GAM")
print(paste0("Peptides not significant in GAM: ", unsign_gam))
```

Look for example at one specific peptide: 

```{r}
p_values_lr[["GLM44"]]["ATVTPSPVK",]
p_values_lr[["GAM"]]["ATVTPSPVK",]
print_pvalues_msqrob(p_values_msqrob, "ATVTPSPVK")

plotlist <- fit_all_and_plot_intensities("ATVTPSPVK", sce)
combined <- wrap_plots(plotlist) & theme(legend.position = "right")
combined + plot_layout(ncol = 2, guides = "collect") # 1200x700 or 1200x1000
```

Study the peptides for which the GAM model has a lack of fit:

```{r}
bad_fit_gam <- get_significant_peptides_per_model(p_values_dev, "GAM")
print(paste0("Peptides with a lack of fit density in GAM: ", bad_fit_gam))

plotlist <- list()
for (peptide in bad_fit_gam[1:9]) {
  plotlist[[peptide]] <- fit_and_plot_intensities(peptide, sce, c("intensities"), "gam",
                                                "as.factor(SampleType)") + ggtitle(peptide)
}

combined <- wrap_plots(plotlist) & theme(legend.position = "bottom")
combined + plot_layout(ncol = 3, nrow = 3, guides = "collect")

table <- data.frame()
for (peptide in bad_fit_gam) {
  table[peptide, "GAM P"] <- p_values_dev[["GAM"]][peptide,"P-value"]
  table[peptide, "GAM AdjP"] <- p_values_dev[["GAM"]][peptide,"Adj-P-value"]
}
table
table%>%xtable(digits = -3)
```
